Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-06-17 Multiresolution_framework/ltx/manuscript.tex
minor edits across paper

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts, such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}), are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}),
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r},
\end{align}
to give the state equation
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1} =\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t,
\end{equation} 
with the parameterized transition matrix
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{equation} 
The problem of joint state and parameter estimation for the state-space model given by Eq.~\eqref{eq:ReducedObservationEquation} and Eq.~\eqref{eq:StateEquation}  can now be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterized  by $\boldsymbol\theta$, known as the likelihood function. This dual estimation problem cannot be solved separately, as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known EM algorithm \citep{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. The EM algorithm is applicable  in two main situations:  first is when the observations are incomplete due to a faulty observation process or existing limitations and the second is when the direct maximization of the likelihood function is difficult but can be simplified by assuming the existence of an additional data referred to as hidden data \citep{Bilmes1998}, the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \citep{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

Both the E and the M steps of the EM algorithm find increasingly tighter lower bounds on the likelihood of the parameters so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For state-space models, the E-step corresponds to the smoothing problem \citep{Haykin2001}, which can be solved using the Rauch Tung Streibel (RTS) smoother \cite{RAUCH1965}. For the M-step, a lower bound on the likelihood function is constructed, where the steps needed
to compute the required quantities are very dependent on the particular application. This will be described in detail for our problem.

@@

where $[\cdot]_{p,q}$ denotes the $\left(p,q\right)$-element of the matrix, and $ [\cdot]^{p,q}$ denotes the $\left(p,q\right)$-block of the block matrix and where 

@@

For each of the experiments, the estimation was applied to the final 900~ms, allowing the forward model's dynamics to stabilize from the initial conditions. The initial state, parameters and the support of the connectivity kernel were unknown to the estimator. The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance function was set to $\eta(r-r') = \sigma_d\times\varphi_{4;3,2}(r-r')$ (B-spline, $j=3$) with $\sigma_d=0.53$.  The firing rate slope, $\varsigma$ was set to $0.56~mV^{-1}$ \citep{Wendling2005}. The observation kernel was modeled using a B-spline function, with a width of 0.08~mm at half the maximum amplitude where the distance between adjacent sensors was $0.05$~mm, resulting into $n_y = 161$ observations. The choice of B-spline function to model the observation kernel is justified by the experiment performed in \citep{Freestone2011}. The spacing and bandwidth of the sensors allowed the full spatial bandwidth of the field to be observed.  With this number of observations, the estimation problem was well-posed, i.e. the number of states, $n_x < n_y$ at each level of the decomposition. This is because the scaling functions are orthogonal to the wavelets at the same and higher levels of decomposition, and  the wavelets are orthogonal to each other across different scales (see Eqs.~\eqref{eq:MRA-PsiPsiOrthogonality} and \eqref{eq:MRA-PhiPsiOrthogonality}).

@@

\section{Acknowledgements}
The research reported herein was partly supported by the Brain Network Recovery Group through the James S. McDonnell Foundation and the FP7-ICT BrainScales.

>>>>

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer physiological parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts, such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}), are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}),
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r},
\end{align}
to give the state equation
\begin{equation}\label{eq:StateEquation}
 \mathbf x_{t+1} =\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t,
\end{equation} 
with the parameterized transition matrix
\begin{equation}\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta)= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{equation} 
The problem of joint state and parameter estimation for the state-space model given by Eq.~\eqref{eq:ReducedObservationEquation} and Eq.~\eqref{eq:StateEquation}  can now be formulated as 
\begin{equation}
	{\hat{\mathbf X},\hat{\boldsymbol\theta}}=\arg\max_{\mathbf X,\boldsymbol\theta}p(\mathbf Y;\boldsymbol\theta),
 \end{equation}  
where $\mathbf X$ and $\mathbf Y$ denote the entire sequences of the states and observations respectively, $\boldsymbol\theta$ is the parameter set and $p(\mathbf Y;\boldsymbol\theta)$ denotes the probability density function of the observations parameterized  by $\boldsymbol\theta$, known as the likelihood function. This dual estimation problem cannot be solved separately, as for the state estimation the knowledge of the system's parameters is required and for the parameter estimation the states of the system need to be known. One approach is to iteratively estimate the states and parameters of the system and monitor the convergence of the algorithm. A natural framework to perform this is the well known EM algorithm \citep{Dempster1977,Shumway2000} to infer both the parameter set and the states from the observations corresponding to the connectivity kernel, the neural field and electrophysiological data in our application. The EM algorithm is applicable  in two main situations:  first is when the observations are incomplete due to a faulty observation process or existing limitations, and the second is when the direct maximization of the likelihood function is difficult but can be simplified by assuming the existence of an additional data referred to as hidden data \citep{Bilmes1998}, which are the weights of the field basis functions in this case. The EM algorithm, when used in the latter context \citep{Dewar2009}, yields the maximum likelihood kernel estimate, i.e.
\begin{equation}
	\boldsymbol\theta_{\text{ML}}=\arg\max_{\boldsymbol\theta}~p(\mathbf Y;\boldsymbol\theta),
 \end{equation}   
and the posterior distribution of the field over time.

Both the E and the M steps of the EM algorithm find increasingly tighter lower bounds on the likelihood of the parameters so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For state-space models, the E-step corresponds to the smoothing problem \citep{Haykin2001}, which can be solved using the Rauch Tung Streibel (RTS) smoother \citep{RAUCH1965}. For the M-step, a lower bound on the likelihood function is constructed, where the steps needed
to compute the required quantities are very dependent on the particular application. This will be described in detail for our problem.

@@

and $[\cdot]_{p,q}$ denotes the $\left(p,q\right)$-element of the matrix, $ [\cdot]^{p,q}$ denotes the $\left(p,q\right)$-block of the block matrix and where 

@@

For each of the experiments, the estimation was applied to the final 900~ms, allowing the forward model's dynamics to stabilize from the initial conditions. The initial state, parameters and the support of the connectivity kernel were unknown to the estimator. The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance function was set to $\eta(r-r') = \sigma_d\times\varphi_{4;3,2}(r-r')$ (B-spline, $j=3$) with $\sigma_d=0.53$.  The firing rate slope, $\varsigma$ was set to $0.56~mV^{-1}$ \citep{Wendling2005}. The observation kernel was modeled using a B-spline function, with a width of 0.08~mm at half the maximum amplitude where the distance between adjacent sensors was $0.05$~mm, resulting into $n_y = 161$ observations. The choice of B-spline function to model the observation kernel is justified by the experiment performed in \citet{Freestone2011}. The spacing and bandwidth of the sensors allowed the full spatial bandwidth of the field to be observed.  With this number of observations, the estimation problem was well-posed, i.e. the number of states, $n_x < n_y$ at each level of the decomposition. This is because the scaling functions are orthogonal to the wavelets at the same and higher levels of decomposition, and  the wavelets are orthogonal to each other across different scales (see Eqs.~\eqref{eq:MRA-PsiPsiOrthogonality} and \eqref{eq:MRA-PhiPsiOrthogonality}).

@@

\section{Acknowledgements}
The research reported herein was partly supported by the Brain Network Recovery Group through the James S. McDonnell Foundation and the FP7-ICT BrainScales. This research was also partly supported by the Australian Research Council (Linkage Project LP100200571).
